Infant Birth Weight in Brazil

### 2024-11-22 products/manuscript/Manuscript_Final_SSM.qmd
Final Changes for RR
Created DiB; updates for all changes

In 1888, Brazil became the last country in the Americas to abolish slavery. Historians have outlined the racialized health disparities of people of African descent in the post-abolition period. Epidemiologists have shown that health disparities continue to mirror patterns from over a century ago. This cross-sectional analysis quantifies health disparities in the post-abolition period using infant birth weight in one of Brazil's first public maternity hospitals. It relies on hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras in Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality. African ancestry was correlated with lower birth weights. Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value = 0.002 [95% CI -138, -32]) than infants born to Euro-descendant women. Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value = 0.019 [95% CI -131, -12]) than infants born to White women. The findings were likely the consequence of slavery’s legacy, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for people of African descent -- and possible epigenetic effects from the lived experience of racism. The findings are consistent with current-day research on racialized health disparities in Brazil.

@@

This study analyzes infant birth weight as a proxy for general maternal and fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras, a public maternity hospital in Rio de Janeiro, Brazil between 1922 and 1926. It aims to provide basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil, substantiating the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.

Birth weight and its relationship to infant health have been a focus of public health research for over a century [@hughes2017]. In epidemiology, both low birth weight (LBW) and preterm birth (PTB), or infants born <37 weeks' gestation, within and across racialized communities have garnered attention [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001]. LBW can be a function of PTB, of fetal growth restriction (FGR, previously known as intra-uterine growth restriction or IUGR, which can include but is not limited to small-for-gestational-age or SGA infants -- those born at less than the 10th percentile), or of both [@wilcox2001; @wilcox2024]. Nonetheless, PTB is an important determinant of both LBW and lower birth weight in general [@maruyama2020].

Multiple social, economic, biological, and genetic factors influence birth weight, which is also partially conditioned by maternal nutrition, health, and health care utilization during pregnancy [@braveman2021a; @david1997; @ward1993]. Recent genetic studies have detailed the specific alleles through which both the maternal and fetal genomes contribute to birth weight and gestational length, while acknowledging the continued importance of non-genetic factors [@juliusdottir2021; @sole-navais2023; @warrington2019]. Given that these studies are based on well-off populations in countries with low neonatal and infant mortality rates, they might overestimate genetic impact on birth weight for historical studies of impoverished populations.

Although not a causal mechanism, LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019]. Originally, epidemiologists hypothesized an association between at-term LBW and increased risk for cardiovascular diseases and other chronic physical health conditions over the life course [@barker2004]. Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest long-term trends associated with "fetal programming" [@maruyama2020; @warrington2019]. Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure. Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.

Historians, using hospital records, have shown how birth weight can be an indicator for the often unmeasurable nutritional status of mothers of different social classes in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @ward1993]. Historically, women from more privileged groups had heavier infants than women from less advantaged groups [@stein1984; @ward1993].

The differences between and within groups both today and in the past have given rise to much research stratified by racial classification. Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. Thus, racism and racist policies also can influence birth outcomes. Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019]. Non-White mothers also had infants with lower mean birth weights than their White counterparts [@leal2006]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997]. Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].

Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@jasienska2009; @owens2019; @martins2024; @steckel1986]. Studies in the human sciences have hypothesized an intergenerational effect of poor health conditions, including undernutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024]. In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal-infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].

Brazil imported over five million enslaved Africans during the nearly four centuries of the existence of chattel slavery [@slavevoyagesconsortium2021], compared to around 300,000 for the United States [@slavevoyagesconsortium2021]. Mortality rates that outstripped fertility rates in Brazil, in addition to higher rates of individual manumission -- the freeing of an enslaved person, can help explain Brazil reliance on the transatlantic slave trade in comparison with the United States, where endogenous growth among the enslaved population occurred by the early nineteenth century [@hacker2020]. Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere. In the nineteenth century, population-level health was poor for all Brazilians, but especially for enslaved individuals, who experienced epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever as well as undernutrition and violent corporal punishment [@karasch1987]. Excluding pregnant and nursing women, historians have argued that caloric intake for the enslaved population was probably sufficient in terms of quantity, yet it lacked key nutrients, and thus the enslaved suffered from nutritional deficiencies, including insufficient thiamine, which was associated with infant mortality [@karasch1987; @kiple1989]. Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes including neonatal death [@gr2024; @kareem2023].

In the aftermath of abolition, no state-run efforts to incorporate formerly enslaved, African-descended peoples into economic, civic, or political life occurred [@fischer2008]. Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, had lower literacy levels, and had worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].

Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019]. The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe. Improved sanitation measures helped stem infectious disease outbreaks [@benchimol1999; @hochman2013]. Infant mortality rates declined from the nineteenth century, although the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@adamo1998; @roth2020; @loudon1992; @woods2009].

Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided healthcare free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020]. Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe.

In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon. In our current postgenomic era, race as a biological category has been rejected, yet the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking in epigenetics remains very real [@meloni2022]. Although differences between racialized groups existed, they were not immutable or due to genetic differences based on race. The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].

@@

This analysis draws on a unique dataset of 2,845 recorded clinical visits, most of them deliveries, to Maternidade Laranjeiras between 1922 and 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926).

In the initial months of publication, the reports included more complete information, including labor time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the obstetrician [@roth2020; @roth2020a]. Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021]. Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.

I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), and the mother’s reproductive history.

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil. Between January 2012 and July 2013, I manually digitized the journal by photographing each volume. Then, between January and August 2017, I manually input the data into Excel from the digital reproductions. I then converted this into a .csv file for upload into R version 4.4.1 (2024-06-14).

The project's Github repository, particularly the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting. All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available online.

@@

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only, in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. In a secondary analysis (see Supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022].

@@

*Maternal skin color*. The original records categorized maternal skin color as White (*branca*), the reference group; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and mixed-race patients into one category, `Afro-Descent`, and White patients into a `Euro-Descent` category.

Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum rather than as a Black-White binary [@travassos2004]. Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category. This method also is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

@@

*Birth*. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention and a vaginal delivery; interventionist, indicating medium medical intervention using forceps and a vaginal delivery; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries), coded as operatory or natural by hospital records, as interventionist. I classified spontaneous abortion as a separate category.

*Infant length*. Infant length was recorded in centimeters in the original records.

*Maternal outcome*. Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintained. I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.

*Fetal outcome*. I recorded fetal outcome as live birth or stillbirth. For the statistical models, I excluded both spontaneous abortions (recorded in the birth variable) and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births in the original data, which is standard methodology for looking at infant birth weight today [@nyarko2013; @wehby2015a; @wehby2017a]. The hospital records recorded 20 neonatal deaths, which I included in the live birth category. These numbers could be higher, as the hospital only recorded deaths within the first week of life, when the infant was still hospitalized.

@@

where maternal deaths (MD) are divided by 10,000 live births (LB). In current studies, MD are divided by 100,000 LB, but I use 10,000 in the denominator, which is in line with historical studies given the higher rates of deaths in the past and reporting uncertainties [@loudon1992].

@@

where the total number of stillbirths (SB) is divided by 1000 total births (TB).

Finally, I calculated the sex ratio at birth (SRaB).

The SRaB is:

$$SRaB = \frac{M}{F} \times 100$$

where the total number of live male births (M) is divided by 100 live female births (F) in a given period.

@@

I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight. Due to the lack of comprehensive information on gestational age, I did not consider birth weight in reference to gestational age. I discuss this constraint in the discussion below.

I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race to see if there were differential outcomes for Black and mixed-race women. The independent variables include maternal age, maternal skin color, gestational status, and maternal nationality.

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 2.

@@

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 3.

Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts of all skin colors [@franken2019; @ball2020]. Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 (modified ancestry) in two steps. First, I created a new dichotomous variable, `Brazilian`/`Non-Brazilian` and ran a simple linear analysis with birth weight as the outcome and maternal nationality as the independent variable. Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value = 0.001 [95% CI 40, 161]). Given that all women of color in the data set were Brazilian, collinearity exists. Second, I dropped all non-Brazilians and ran the original simple and multiple regressions from Models 1 and 2. Results for the sub-sample analysis are included in Table 2.

@@

Table 1 provides summary statistics for the dataset. In the hospital records, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*). First-time mothers (nulliparous) comprised 41%. Of all women, 84% were Brazilian. Of the nearly 16% immigrant patients, 15% were from Europe.

Of all reproductive outcomes, excluding spontaneous abortion, 91% were natural deliveries. The remaining 9% were interventionist or operative deliveries. For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94.1%. The SBR was 82.9%. The sex ratio was 1.2: there were 120 male live births per 100 female live births.

@@

This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in the early twentieth century, a hospital that remains in operation today. It provides quantitative evidence of racialized disparities in maternal-infant health in early twentieth-century Brazil. These findings are thus useful to compare to present-day research to assess the extent of progress in improved outcomes over time. They also complement descriptive historical studies and provide a base for qualitative research to analyze individual experience in the past.

>>>>

In 1888, Brazil became the last country in the Americas to abolish slavery. Historians have outlined the racialized health disparities of people of African descent in the post-abolition period. Epidemiologists have shown that twenty-first-century health disparities continue to mirror patterns from over a century ago. This cross-sectional analysis quantifies health disparities in a post-abolition maternity hospital using infant birth weight. It relies on hospital records on infants delivered between 1922-1926 (n=2,845) at the Maternidade Laranjeiras in the city of Rio de Janeiro, Brazil to run linear models assessing differences in infant birth weight by maternal skin color, age, number of pregnancies (parity), and nationality. African ancestry was correlated with lower birth weights. Infants born to Afro-descendant women had birth weights estimated to be 84 grams lighter (p-value = 0.002 [95% CI -137, -32]) than infants born to Euro-descendant women. Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value = 0.001 [95% CI -160, -39]) and infants born to mixed-race (*parda*) women had birth weights estimated to be 70 grams lighter (p-value = 0.022 [95% CI -130, -10]) than infants born to White women. The findings were likely the consequence of slavery’s legacy, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for people of African descent. The findings are consistent with current-day research on racialized health disparities in Brazil and demonstrate the importance of historical findings to public health research.

@@

This study analyzes infant birth weight as a proxy for general maternal, fetal, and infant health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras, a public maternity hospital in the city of Rio de Janeiro, Brazil between 1922 and 1926. It provides basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil, substantiating researcher's qualitative findings on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.

Birth weight and its relationship to fetal and infant health have been a focus of public health research for over a century [@hughes2017; @schneider2017]. In epidemiology, both low birth weight (LBW) and preterm birth (PTB), or infants born <37 weeks' gestation, within and across racialized communities have garnered attention [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001]. LBW can be a function of PTB, of fetal growth restriction (FGR, previously known as intra-uterine growth restriction or IUGR, which can include but is not limited to small-for-gestational-age or SGA infants -- those born at less than the 10th percentile), or of both [@wilcox2001; @wilcox2024]. Nonetheless, PTB is an important determinant of both LBW and lower birth weight in general [@maruyama2020].

Multiple social, economic, biological, and genetic factors influence birth weight, which is also partially conditioned by maternal nutrition, health, and health care utilization during pregnancy [@braveman2021a; @david1997; @schneider2017; @ward1993]. Recent genetic studies have detailed the specific alleles through which both the maternal and fetal genomes contribute to birth weight and gestational timing, while acknowledging the continued importance of non-genetic factors [@sole-navais2023; @warrington2019].

Although not a causal mechanism, LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019]. Originally, epidemiologists hypothesized an association between at-term LBW and increased risk for cardiovascular diseases and other chronic physical health conditions over the life course [@barker2004]. Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest long-term trends associated with "fetal programming." Warrington et al [-@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure. Maruyama and Heinesen [-@maruyama2020] have found that although lower birth weight is associated with both infant mortality and other health issues in the neonatal period, its association with long-term health issues declines over time.

Historians, using hospital records, have shown how birth weight can be an indicator for the often-unmeasurable nutritional status of mothers in the past [@costa1998; @costa2004; @goldin1989; @steckel1986; @ward1993]. Historically, women from more privileged groups had heavier infants than less-advantaged women [@stein1984; @ward1993].

The differences between and within groups both today and in the past have prompted racially-stratified research. While race is a social construct and thus biological fiction, the social consequences of racism influence health and disease patterns, including birth outcomes, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019]. Non-White mothers also have infants with lower mean birth weights than their White counterparts [@leal2006]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].

Scholars across disciplines have argued that racialized health disparities in countries with histories of race-based chattel slavery are longstanding and tied to the unequal and violent social relations produced under that institution [@jasienska2009; @owens2019; @steckel1986]. Studies in the human sciences have hypothesized an intergenerational effect of poor health conditions, including undernutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020]. Historians have emphasized how race and socioeconomic status (SES) were deeply intertwined in post-abolition Brazil and thus disparities in literacy and access to education likely contributed to health gaps [@adamo1998; @ball2020; @franken2025]. In Brazil, most historical studies of health under slavery and its aftermath remain descriptive and have not tested quantitative associations between the institution's legacies, including both race-based socioeconomic inequality and the deleterious effects of racism, and maternal, fetal, and infant health outcomes [@adamo1998; @otovo2016; @roth2020; @telles2022].

Brazil imported over five million enslaved Africans over nearly four centuries [@slavevoyagesconsortium2021]. Mortality rates that surpassed fertility rates, in addition to higher rates of individual manumission -- the freeing of an enslaved person -- helps explain Brazil's reliance on the transatlantic slave trade. Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere. In the nineteenth century, population-level health was poor for all Brazilians, but especially for enslaved individuals, who experienced epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever as well as undernutrition and violent punishment [@karasch1987]. Excluding pregnant and nursing women, historians have argued that caloric intake for the enslaved population was sufficient in terms of quantity, yet it lacked key nutrients. Thus, the enslaved suffered from nutritional deficiencies, including thiamine, which was associated with infant mortality [@karasch1987; @kiple1989]. Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes [@kareem2023].

In the aftermath of abolition, no state-run efforts to incorporate formerly enslaved peoples into economic, civic, or political life occurred [@fischer2008]. Further, the state supported a policy of whitening in which elites believed that both racial mixture and White immigration would whiten the population (Ball, 2020; Stepan, 1991). Scholars have shown how Black and mixed-race Brazilians were incarcerated at higher rates, and had lower literacy levels, fewer employment opportunities, and worse health outcomes than their White and White immigrant counterparts [@adamo1998; @ball2020; @fischer2008; @otovo2016].

Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in urban centers in Brazil's southeastern and southern regions, including Rio de Janeiro [@franken2025]. The early twentieth century was a period of advancement in both clinical care and public health in Brazil and globally. In cities, improved sanitation measures stemmed infectious disease outbreaks [@benchimol1999; @hochman2013]. Infant mortality rates declined from the nineteenth century, although the major advancements including blood transfusions and antibiotics that resulted in sustained improvements in maternal and infant mortality and stillbirth rates only began in the late 1930s [@adamo1998; @roth2020; @loudon1992; @woods2009]. Further, these advancements were concentrated in cities, and rural-urban and regional divides in maternal-infant health outcomes developed across Brazil that only decreased in the new Millennium [@scheper-hughes1992; @victora2011].

A rural-urban and regional divide was also noticeable in fertility patterns. Women in the city of Rio de Janeiro started their fertility transition in the 1920s. In the early 1930s, the total fertility rate (TFR), which summarizes fertility rates across all women of reproductive age in a particular period, was less than 5 in Rio de Janeiro compared to 6 for São Paulo state [@goncalves2019]. By 1940, TFR for Rio de Janeiro had reached 3.8, compared to 4.7 for the city of São Paulo and 6.4 for Brazil [@ibge1952].

Maternidade Laranjeiras opened in 1904 and became the obstetrical teaching hospital at Rio de Janeiro’s medical school in 1918. In the mid-1920s, it was one of Rio de Janeiro's three maternity hospitals for a total of 164 beds in a city of nearly 1.2 million inhabitants [@carvalho1924; @penteado1924; @roth2020].

Laranjeiras provided healthcare free of charge. The hospital was built in a neighborhood near weaving and textile factories that employed women workers. Scholars have found that most of its patients were neighborhood working-class women who were poor but not indigent. Most patients were women of color defined as either Black or mixed race, and most White patients were European immigrants [@barreto2016; @roth2020]. Because it was the medical school's teaching hospital, it also attracted difficult labors and deliveries, with women of all classes referred by a physician or midwife who initially attended the birth at home [@pintofilho1923]. Although other maternity hospitals in the 1920s had ambulatory systems, which provided home-based prenatal and labor and delivery care, Laranjeiras did not [@roth2020].

Because most of the hospital's patients were working-class women from the surrounding neighborhood, its clientele was relatively representative of the city's population, which was majority lower SES. But the hospital probably had a disproportionate number of difficult births given that medical practitioners referred obstructed labors to Laranjeiras. Further, in 1920s Rio de Janeiro, most mothers of all SES gave birth at home. At the turn of the century, around 99% of births were home-based. By the mid-1930s, around 80% of births occurred at home. For normal deliveries, wealthier women were least likely to give birth in a hospital [@roth2020].

In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Rio de Janeiro, my aim is not to reify race as a genetic or epigenetic phenomenon [@meloni2022]. Although differences between racialized groups existed, they were not immutable or due to race-based genetic differences. The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not part of permanently "damaged" communities (socially or biologically) [@tuck2009] but rather were living the consequences of slavery and socioeconomic and racial inequality.

@@

This analysis draws on a unique dataset of 2,845 recorded clinical visits, majority births, to Maternidade Laranjeiras in the city of Rio de Janeiro, Brazil between 1922 and 1926. I extracted data from the country's premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO), which, between June 1922 and May 1926, published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926).

I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery (natural, interventionist, operative), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth, live birth, or neonatal death), twin status (singleton or twin), and maternal reproductive history. Hospital clinicians erratically recorded civil status and gestational age, and thus I excluded them.

The journal is held at the Biblioteca Nacional (BN), the Maternidade Escola, Rio de Janeiro (ME-UFRJ), and the Biblioteca CB/A-Biomédica A, Universidade Estadual do Rio de Janeiro (UERJ), all in Rio de Janeiro, Brazil. Between January 2012 and July 2013, I photographed each volume. Then, between January and August 2017, I manually input the data into Excel from the digital reproductions. I then converted this into a .csv file for upload into R version 4.4.1 (2024-06-14).

The project's Github repository, particularly the `Supplemental_Materials` file, includes summary information on data processing, exploratory analysis, and model fitting. All materials, including the data set and code, are open access and reproducible.

@@

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. In `Supplemental_Materials`, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@who2022].

@@

*Maternal skin color*. The original records categorized maternal skin color as White (*branca*), the reference group; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. I also created a new skin-color dummy variable, combining Black and mixed-race patients as `Afro-Descent` and White patients as `Euro-Descent`.

Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum [@travassos2004]. Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category. This method also aligns with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

@@

*Birth*. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention and a vaginal delivery; interventionist, indicating medium medical intervention using forceps and a vaginal delivery; and operatory, indicating a cesarean section or embryotomy. I recategorized external manipulations including version and Mauriceau maneuver, coded as operatory or natural by hospital records, as interventionist. I classified spontaneous abortion as a separate category.

*Infant length*. Hospital records included infant length in centimeters.

*Maternal outcome*. Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintained. I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality.

*Fetal outcome*. I recorded fetal outcome as live birth or stillbirth. For the statistical models, I excluded both spontaneous abortions (recorded in the birth variable) and any recorded live births weighing \<500 grams to avoid the inclusion of possible stillbirths that were misrecorded as live births in the original data [@nyarko2013; @wehby2015a]. The hospital records listed 20 neonatal deaths, which I included as live births. These numbers could have been higher, as the hospital only recorded deaths within the first week of life, when the infant was still hospitalized. I ran all linear models on singleton births only, excluding twins [@david1997; @leal2006; @nyarko2013; @pollock2021; @silva1998; @silveira2019; @vilanova2019; @wehby2015a].

@@

where maternal deaths (MD) are divided by 10,000 live births (LB). Today, MD are divided by 100,000 live births, but I use 10,000 live births in the denominator given the higher rates of deaths in the past and reporting uncertainties [@loudon1992]. The MMR measures obstetric risk: the risk of death once a person is pregnant [@who2006].

@@

where the total number of stillbirths (SB) is divided by 1,000 total births (TB).

@@

I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight. I ran four linear models: two in which I categorized maternal skin color into two ancestral groups: White (Euro-descent), the reference group, and non-White (Afro-descent), and two in which I categorized maternal skin color according to the hospital records' original racial classifications: White (the reference group), Black, and mixed race to see if there were differential outcomes for Black and mixed-race women. The independent variables include maternal age, skin color, parity, and nationality, and year fixed effects.

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Par` is parity (nulliparous or multiparous), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term. Results for both models are displayed in Table 2.

@@

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Par} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \text{YrFE} + \epsilon$$ whereas `BW` is birth weight in grams, $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `MC` is maternal skin color (White, Black, and mixed race), `Par` is parity (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, `YrFE` is year fixed effect, and $\epsilon$ is the error term. Results for both models are displayed in Table 3.

Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts of all skin colors [@ball2020; @franken2025]. Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 (modified ancestry) in two steps. First, I created a new dichotomous variable, `Brazilian`/`Non-Brazilian` and ran a simple linear analysis with birth weight as the outcome and maternal nationality as the independent variable. Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value = 0.001 [95% CI 40, 161]). Given that all women of color in the data set were Brazilian, collinearity exists. Second, I dropped all non-Brazilians and ran the original regressions from Models 1 and 2. Results for the sub-sample analysis are included in Table 2.

@@

Table 1 provides summary statistics for the dataset. Of hospital patients, 58% were of African descent (defined as *preta* or *parda*) and 42% were of European descent (*branca*). An average of census data for the city's female population's racial makeup between 1890 and 1940, years for which data exist, show that White women comprised 63%, mixed-race women 24%, and Black women 14% (with rounding errors) [@roth2020]. Thus, at Laranjeiras, women of color were over-represented among hospital patients. First-time mothers (nulliparous), comprised 41%. Of all women, 84% were Brazilian. Of the nearly 16% immigrant patients, 15% were European.

Of all reproductive outcomes, excluding spontaneous abortion, 91% were natural deliveries. The remaining 9% were interventionist or operative deliveries. For patients who went to the clinic to deliver their infant (excluding spontaneous abortion), 23 died, for an MMR of 94 per 10,000 live births. The SBR was 83 stillbirths per 1,000 live births.

@@

This study contributes heretofore unpublished hospital records from one of Brazil's most important public maternity hospitals in a major early twentieth-century city, a hospital that remains in operation today. It provides quantitative evidence of racialized disparities in maternal, fetal, and infant health. These findings are thus useful to compare to present-day research to assess the extent of progress in outcomes. They also provide direct evidence of historical racial disparities in Brazil, rare for this period, and thus complement descriptive historical studies.
